---
title: "Multivar Stat Distr in R"
author: "alexmerk"
date: "November 30, 2018"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
#install.packages("pacman", repos="http://cran.us.r-project.org")
library(pacman)
pacman::p_load(ggplot2,
               GGally,
               mvtnorm,
               sn,
               MVN,
               scatterplot3d,
               factoextra,
               corrplot)
```

## 1 Reading and plotting multivariate data
### 1.1 Reading MV data
```{r}
# Reading the data sets from the URL (a Github repo)
ais <- read.csv(file="http://raw.githubusercontent.com/alexmerk/xcourse/master/mva/ais.csv")
wine <- read.csv(file="http://raw.githubusercontent.com/alexmerk/xcourse/master/mva/wine.csv")

# Change the wine type variable to a factor
wine$Type <- as.factor(wine$Type)
```

### 1.2 Mean vector and variance-covariance matrix
#### Several commands to calculate means (also by group)
```{r}
# Calculate the mean of the Alcohol, Malic, Ash, and Alcalinity variables 
colMeans(wine[,2:5])

# Calculate the mean of the variables by wine type
by(wine[,2:5], INDICES=wine$Type, FUN=colMeans)

# Or using the tidyverse/purrr
#wine %>% map_dbl(mean, na.rm=TRUE)
```

#### Variance-covariance and correlation matrices
```{r}
# Calculate the variance-covariance matrix of the variables Alcohol, Malic, Ash, Alcalinity
var.wine <- var(wine[,2:5])

# Round the matrix values to two decimal places 
round(var.wine, 2)

# Calculate the correlation matrix 
cor.wine <- cor(wine[,2:5])

# Round the matrix to two decimal places 
round(cor.wine,2)

# Plot the correlations 
corrplot(cor.wine, method = "ellipse")
```

### 1.3 Plotting MV data
#### Using R base & ggplot2 graphics
```{r}
# Scatter plot matrix using the base R plot function
pairs(wine[,2:5])

# Produce a matrix of plots for the first four variables 
wine.gg <- ggpairs(wine, columns=2:5, mapping=aes(color=Type))
wine.gg

# Plot three variables 
scatterplot3d(wine[, c(2,3,5)], color=wine$Type)
```

## 2 Multivariate Normal Distribution
### 2.1 MND: Definition and sampling
#### Samples from MVND=MND
```{r}
# Given following parameters:
mu.sim <- c(2,-2)
sigma.sim <- matrix(c(9,5,5,4), nrow=2)

# Generate 100 bivariate normal samples
multnorm.sample <- rmvnorm(100, mean=mu.sim, sigma=sigma.sim)

# View the first 6 samples
head(multnorm.sample)

# Scatterplot of the bivariate samples 
#plot(multnorm.sample[,1],multnorm.sample[,2])
#Or shorter:
    plot(multnorm.sample)
```

### 2.2 Density of MND
#### Calculating the density of MND
```{r}
# Calculate density
multnorm.dens <- dmvnorm(multnorm.sample,
            mean =mu.sim, sigma = sigma.sim)

# Create scatter plot of density heights 
scatterplot3d(cbind(multnorm.sample,multnorm.dens),    
               color="blue", pch="", type = "h",             
                              xlab = "x", ylab = "y", zlab = "density")
```

#### Calculating the density over a grid
```{r}
# Creating a 40 by 40 grid
mvals <- expand.grid(seq(-5, 10, length.out = 40), seq(-8, 4, length.out = 40))

# Calculate density over the specified grid
mvds <- dmvnorm(as.matrix(mvals),mean=mu.sim, sigma=sigma.sim)
matrix_mvds <-  matrix(mvds, nrow = 40)

# Create a perspective plot
persp(matrix_mvds, theta = 80, phi = 30, expand = 0.6, shade = 0.2, col = "lightblue", xlab = "x", ylab = "y", zlab = "dens")
```

### 2.3 Cumulative Distribution and Inverse CDF
#### Calculate Volume for a biv standard ND and for a biv ND with given Mu, Sigma
```{r}
# Volume under a bivariate standard normal
pmvnorm(lower = c(-1,-1), upper = c(1,1))

# Volume under specified mean and variance-covariance matrix
mu.sim <- c(2,-2)
sigma.sim <- matrix (c(9,5,5,4), nrow=2)
pmvnorm(lower=c(-5,5), upper=c(-5,5), mean = mu.sim, sigma = sigma.sim)
```

#### Calculating probability contours using qmvnorm
```{r}
# Probability contours for a standard bivariate normal
qmvnorm(0.9, tail = "both", sigma = diag(2))

# Probability contours for a bivariate normal 
qmvnorm(0.95, tail="both", mean=mu.sim, sigma=sigma.sim)
```

### 2.4 Checking normality of MV data
#### Graphical tests for MV normality
```{r}
# We need the package MVN

# Test sample normality
qqnorm(multnorm.sample[,1])
qqline(multnorm.sample[,1])

# Create qqnorm plot 
#uniPlot(wine[,2:5], type = "qqplot")
## Package changed, use other options

# mardiaTest qqplot 
wine.mvntest <- mvn(wine[,2:5], mvnTest="mardia", multivariatePlot = "qq")
```

#### Numerical tests for multivariate normality
```{r}
# Use mardiaTest
mvn(multnorm.sample, mvnTest="mardia")

# Use hzTest
mvn(wine[, 2:5], mvnTest="hz")
```

## 3 Other MV distributions

### 3.1 Other common MV distributions
```{r}
# Specify delta and sigma
delta <- c(1, 2, -5) 
sigma <- matrix(c(1, 1, 0, 1, 2, 0, 0, 0, 5), 3, 3)

# Generate samples                  
t.sample <- rmvt(n = 2000, delta = delta, sigma = sigma, df = 4)
```

#### Generate samples from MV t-distribution
```{r}
# Generate the t-samples 
multt.sample <- rmvt(200,df=5, delta=mu.sim, sigma=sigma.sim)

# Print the first 6 samples
head(multt.sample)

# Check multivariate normality
mvn(multt.sample, mvnTest="mardia", multivariatePlot = "qq")
```

### 3.2 Density and cumulative density for MV-t
```{r}
x <- seq(-3, 6, by = 1); y <- seq(-3, 6, by = 1)
d <- expand.grid(x = x, y = x)                   
del1 <- c(1, 2); sig1 <- matrix(c(1, .5, .5, 2), 2)
dens <- dmvt(as.matrix(d), delta = del1, sigma = sig1, df = 10, log = FALSE)
scatterplot3d(cbind(d, dens), type = "h", zlab = "density")
```

#### Density of MV t-distribution
```{r}
# Calculate densities
multt.dens <- dmvt(multt.sample, delta=mu.sim, sigma=sigma.sim, df=5) 

# Plot 3D heights of densities
scatterplot3d(cbind(multt.sample, multt.dens),     
                 color = "blue", pch = "", type = "h",             
                              xlab = "x", ylab = "y", zlab = "density")
```

#### Cumulative distributions and quantiles of t
```{r}
# Calculate the volume under the specified t-distribution
pmvt(lower=c(-5,-5), upper=c(5,5), df=5, delta=mu.sim, sigma=sigma.sim)

# Calculate the equal probability contour
qmvt(0.90, tail="both", df=5, delta=mu.sim, sigma=diag(2))
```

### 3.3 Multivariate skewed distributions
#### Drawing samples from skew distributions
```{r}
# Preloading the parameters
mu.sim <- c(2,-2)
sigma.sim <- matrix(c(9,5,5,4), 2, 2)

# Generate the skew-normal samples 
skewnorm.sample <- rmsn(100, xi=mu.sim, Omega=sigma.sim, alpha=c(4,-4)) 
colnames(skewnorm.sample) <- c("x1","x2")
    
# Print first six samples
head(skewnorm.sample)

# Generate the skew-t samples 
skewt.sample <- rmst(n = 100, xi = mu.sim, Omega = sigma.sim, alpha = c(4, -4), nu=5)


# Print first six samples
head(skewt.sample)
```

#### Plotting and testing of skewed-densities
```{r}
# Contour plot for skew-normal sample
gg <- ggplot(data=as.data.frame(skewnorm.sample), aes(x=x1,y=x2)) + geom_point() + geom_density_2d()
gg

# Normality test for skew-normal sample
skewnorm.Test <- mvn(skewnorm.sample, mvnTest="mardia", multivariatePlot = "qq")
# Normality test for skew-t sample
skewt.Test <- mvn(skewt.sample, mvnTest="mardia", multivariatePlot = "qq")
```

#### Parameter estimation for MV skew-normals
```{r}
# Splitting the data set
ais.female <- ais[ais$sex=="female",]
ais.male <- ais[ais$sex=="male",]

# Fit skew-normal parameters
fit.ais <- msn.mle(y = cbind(ais.female$Ht, ais.female$Wt), opt.method = "BFGS")

# Print the skewness parameters
fit.ais$dp$alpha
```

## 4 Principal Component Analysis
### 4.1 Performing PCA
#### Using the princomp function
```{r}
# Calculate PCs
pca.state <- princomp(state.x77, cor=TRUE, scores=TRUE) 

# Plot the PCA object  
plot(pca.state) 

# Print the summary of the PCs
summary(pca.state)  
```

#### Calculating the POV explained & Choosing the number of PCs
```{r}
# Variance explained by each PC
pca.var <- pca.state$sdev^2  

# Proportion of variance explained by each PC
pca.pvar <- pca.var/sum(pca.var) 

# Proportion of variance explained by each principal component
pca.pvar

# Cumulative variance explained plot
plot(cumsum(pca.pvar), xlab = "Principal component", ylab = "Cumulative Proportion of variance explained", ylim = c(0,1), type = 'b')
grid()

# Add a horizontal line
abline(h=0.95, col="blue")

# Alternatively: Choosing the number of PCs using scree plot
stats::screeplot(pca.state, type = "l")
grid()
```

### 4.2 Interpreting PCA attributes
#### Loadings and scores for the PCs using ggplot2
```{r}
# Create dataframe of scores
scores.state <- data.frame(pca.state$scores)

# Have to import:
state.region <- c("Southeast","West","West","Southeast","West","West","Northeast","Southeast",
                  "Southeast","Southeast","West","West","Northcentral","Northcentral",
                  "Northcentral","Northcentral","Southeast","Southeast","Northeast","Southeast",
                  "Northeast", "Northcentral","Northcentral","Southeast","Northcentral","West",
                  "Northcentral","West","Northeast","Northeast","West","Northeast","Southeast",
                  "Northcentral","Northcentral","Southeast","West","Northeast","Northeast",
                  "Southeast","Northcentral","Southeast","Southeast","West","Northeast","Southeast",
                  "West","Southeast","Northcentral","West")

# Plot of scores labeled by state name (showing only the 1st and 2nd component)
ggplot(data = scores.state, aes(x = Comp.1, y = Comp.2, label = rownames(scores.state), color=state.region)) + 
    geom_text(alpha=0.8, size = 3) + 
    ggtitle("PCA of states data")
```

#### Visualizing PCA using the factoextra library
```{r}
#Plot the first two components
pca.score <- pca.state$scores
fviz_pca_ind(pca.state)
# Plot the PC loadings
#fviz_pca_var(pca.state)
# Create a biplot
#fviz_pca_biplot(pca.state)
```

### 4.3 Multi-dimensional scaling
#### MDS in two and three dimensions
```{r}
## MDS in 2D
# Calculate distance 
state.dist <- dist(state.x77)

# Perform multidimensional scaling 
mds.state <- cmdscale(state.dist,k=2) 

# Change mds.state to a dataframe for use in ggplot
mds.state_df <- data.frame(mds.state)

# Plot the representation of the data in two dimensions 
ggplot(data = mds.state_df, aes(x = X1, y = X2, label = rownames(mds.state), color = state.region)) + 
    geom_text(alpha = 0.8, size = 3) 

## MDS in 3D
# Calculate distance 
wine.dist <- dist(wine[,-1])

# Perform multidimensional scaling 
mds.wine <- cmdscale(wine.dist, k=3) 
mds.wine_df <- data.frame(mds.wine)

# Plot the representation of the data in three dimensions 
scatterplot3d(mds.wine_df, color = wine$Type, pch = 19, type = "h", lty.hplot = 2)
```
