---
title: "DataCamp - EDA Case Study"
author: "Aleksandr Merkulov"
date: "December 12, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Installing the packages
library(pacman)
pacman::p_load(dplyr, ggplot2, countrycode, broom, purrr, tidyr)
```

```{r import, cache=FALSE, include=FALSE}
# Importing data
votes <- as_tibble(read.csv(
    #file="https://github.com/alexmerk/courses/blob/master/r/datacamp/eda/votes.csv"))
  file="C:/github/course/r/datacamp/eda/votes.csv"))
descriptions <- as_tibble(read.csv(
    file="C:/github/course/r/datacamp/eda/descriptions.csv"))
```
## Exploratory Data Analysis
The following project was completed as a part of Datacamp's R courses. Below is 
the original project description:  
_Once you've started learning tools for data manipulation and visualization like dplyr and ggplot2, this course gives you a chance to use them in action on a real dataset. You'll explore the historical voting of the United Nations General Assembly, including analyzing differences in voting between countries, across time, and among international issues. In the process you'll gain more practice with the dplyr and ggplot2 packages, learn about the broom package for tidying model output, and experience the kind of start-to-finish exploratory analysis common in data science._  

### 1 Data cleaning and summarizing with dplyr
```{r dplyr}
# Filter for votes that are "yes", "abstain", or "no"
# Add another %>% step to add a year column
votes %>%
  filter(vote <= 3) %>%
  mutate(year=session+1945)

# Load the countrycode package
#library(countrycode)

# Convert country code 100
countrycode(100,"cown","country.name")

# Add a country column within the mutate: votes_processed
votes_processed <- votes %>%
  filter(vote <= 3) %>%
  mutate(year = session + 1945, 
  country=countrycode(ccode,"cown","country.name"))

# Find total and fraction of "yes" votes
votes_processed %>% summarize(
total=n(),
percent_yes=mean(vote==1))

# Summarize by country: by_country
by_country <- votes_processed %>%
  group_by(country) %>%
  summarize(total = n(),
            percent_yes = mean(vote == 1))

# Filter out countries with fewer than 100 votes
by_country %>%
  arrange(percent_yes) %>%
  filter(total>=100)
  
```

### 2 Visualizing data with ggplot2
```{r ggplot2}
# Define by_year
by_year <- votes_processed %>%
  group_by(year) %>%
  summarize(total = n(),
            percent_yes = mean(vote == 1))

# Load the ggplot2 package
#library(ggplot2)

# Create line plot
ggplot(by_year, aes(x=year, y=percent_yes)) +
  geom_line()

# Change to scatter plot and add smoothing curve
ggplot(by_year, aes(year, percent_yes)) +
  geom_point() + geom_smooth()
  
# Group by year and country: by_year_country
by_year_country <- votes_processed %>%
  group_by(year, country) %>%
  summarize(total = n(),
            percent_yes = mean(vote == 1))

# Start with by_year_country dataset
by_year_country <- votes_processed %>%
  group_by(year, country) %>%
  summarize(total = n(),
            percent_yes = mean(vote == 1))

# Print by_year_country
by_year_country

# Create a filtered version: UK_by_year
UK_by_year <- filter(by_year_country, country=="United Kingdom")

# Line plot of percent_yes over time for UK only
ggplot(UK_by_year, aes(x=year,y=percent_yes)) +
  geom_line()

# Vector of four countries to examine
countries <- c("United States", "United Kingdom",
               "France", "India")

# Filter by_year_country: filtered_4_countries
filtered_4_countries <- by_year_country %>%
filter(country %in% countries)

# Line plot of % yes in four countries
ggplot(filtered_4_countries, aes(x=year, y=percent_yes, color=country)) +
  geom_line()

# Vector of six countries to examine
countries <- c("United States", "United Kingdom",
               "France", "Japan", "Brazil", "India")

# Filtered by_year_country: filtered_6_countries
filtered_6_countries <- by_year_country %>% filter (country %in% countries)

# Line plot of % yes over time faceted by country
ggplot(filtered_6_countries, aes(x=year, y=percent_yes)) +
geom_line() +
  facet_wrap(~country)


# Line plot of % yes over time faceted by country and a free y-scale
ggplot(filtered_6_countries, aes(year, percent_yes)) +
  geom_line() +
  facet_wrap(~ country, scales="free_y")

# Add three more countries to this list
countries <- c("United States", "United Kingdom",
               "France", "Japan", "Brazil", "India", 
               "Russian Federation", "Pakistan", "Kazakhstan")

# Filtered by_year_country: filtered_countries
filtered_countries <- by_year_country %>%
  filter(country %in% countries)

# Line plot of % yes over time faceted by country
ggplot(filtered_countries, aes(year, percent_yes)) +
  geom_line() +
  facet_wrap(~ country, scales = "free_y")
```

### 3 Tidy modeling with broom
```{r broom}
# Percentage of yes votes from the US by year: US_by_year
US_by_year <- by_year_country %>%
  filter(country == "United States")

# Print the US_by_year data
US_by_year

# Perform a linear regression of percent_yes by year: US_fit
US_fit <- lm(formula=percent_yes ~ year, data=US_by_year)

# Perform summary() on the US_fit object
summary(US_fit)

# Load the broom package
#library(broom)

# Call the tidy() function on the US_fit object
tidy(US_fit)


# Linear regression of percent_yes by year for US
US_by_year <- by_year_country %>%
  filter(country == "United States")
US_fit <- lm(percent_yes ~ year, US_by_year)

# Fit model for the United Kingdom
UK_by_year <- by_year_country %>%
filter(country=="United Kingdom")
UK_fit <- lm(percent_yes ~ year, UK_by_year)

# Create US_tidied and UK_tidied
US_tidied <- tidy(US_fit)
UK_tidied <- tidy(UK_fit)

# Combine the two tidied models
bind_rows(US_tidied, UK_tidied)

```

#### Nesting the data
```{r nesting}
# Load the tidyr package
#library(tidyr)

# Nest all columns besides country
by_year_country %>% nest(-c(country))

# All countries are nested besides country
nested <- by_year_country %>%
  nest(-country)

# Print the nested data for Brazil
nested$data[nested$country=="Brazil"]

# Unnest the data column to return it to its original form
nested %>% unnest(data)
```

```{r map lm}

# IMPORTANT: Ungroup the data set before nesting it.
by_year_country <- by_year_country %>% ungroup()


# Perform a linear regression on each item in the data column
by_year_country %>%
  nest(-country) %>% mutate(model=map(data, ~ lm(percent_yes ~ year, .)))
  
# Load the broom package
#library(broom)

# Add another mutate that applies tidy() to each model
by_year_country %>%
  nest(-country) %>%
  mutate(model = map(data, ~ lm(percent_yes ~ year, data = .))) %>%
  mutate(tidied = map(model, ~tidy(.)))

# Add one more step that unnests the tidied column
country_coefficients <- by_year_country %>%
  nest(-country) %>%
  mutate(model = map(data, ~ lm(percent_yes ~ year, data = .)),
         tidied = map(model, ~tidy(.))) %>%
         unnest(tidied)

# Print the resulting country_coefficients variable
country_coefficients


### Working with many tidy models ###
# Print the country_coefficients dataset
country_coefficients

# Filter for only the slope terms
filter(country_coefficients, term=="year")

# Filter for only the slope terms
slope_terms <- country_coefficients %>%
  filter(term == "year")

# Add p.adjusted column, then filter
slope_terms %>% mutate (p.adjusted=p.adjust(p.value)) %>%
filter(p.adjusted <0.5)

# Filter by adjusted p-values
filtered_countries <- country_coefficients %>%
  filter(term == "year") %>%
  mutate(p.adjusted = p.adjust(p.value)) %>%
  filter(p.adjusted < .05)

# Sort for the countries increasing most quickly
arrange(filtered_countries, estimate)

# Sort for the countries decreasing most quickly
arrange(filtered_countries, desc(estimate))
```

### 4 Joining data sets
```{r warning=FALSE}
# Load dplyr package
#library(dplyr)

# Print the votes_processed dataset
votes_processed

# Print the descriptions dataset
descriptions

# Join them together based on the "rcid" and "session" columns
votes_joined <- inner_join(votes_processed, descriptions, 
by=c("rcid", "session"))

# Filter for votes related to colonialism
votes_joined %>% filter(co==1)

# Load the ggplot2 package
#library(ggplot2)

# Filter, then summarize by year: US_co_by_year
US_co_by_year <- votes_joined %>% filter(co==1, country=="United States") %>% group_by(year) %>% summarize(percent_yes=mean(vote==1, na.rm=TRUE)) 

# Graph the % of "yes" votes over time
ggplot(US_co_by_year, aes(x=year, y=percent_yes)) +
geom_line()


# Load the tidyr package
#library(tidyr)

# Gather the six me/nu/di/hr/co/ec columns
votes_joined %>%
  gather(topic, has_topic, me:ec)

# Perform gather again, then filter
votes_gathered <- votes_joined %>%
  gather(topic, has_topic, me:ec) %>%
  filter(has_topic == 1)

# Replace the two-letter codes in topic: votes_tidied
votes_tidied <- votes_gathered %>%
  mutate(topic = recode(topic,
                        me = "Palestinian conflict",
                        nu = "Nuclear weapons and nuclear material",
                        di = "Arms control and disarmament",
                        hr = "Human rights",
                        co = "Colonialism",
                        ec = "Economic development"))

# Print votes_tidied
votes_tidied

# Summarize the percentage "yes" per country-year-topic
by_country_year_topic <- votes_tidied %>% group_by(country, year, topic) %>%
summarize(total=n(), percent_yes=mean(vote==1, na.rm=TRUE)) %>% ungroup()

# Print by_country_year_topic
by_country_year_topic

# Load the ggplot2 package
#library(ggplot2)

# Filter by_country_year_topic for just the US
US_by_country_year_topic <- by_country_year_topic %>% filter (country=="United States")

# Plot % yes over time for the US, faceting by topic
ggplot (US_by_country_year_topic, aes (x=year, y=percent_yes)) + geom_line() +
facet_wrap(~topic)

# Print by_country_year_topic
by_country_year_topic

# Fit model on the by_country_year_topic dataset
country_topic_coefficients <- by_country_year_topic %>%
nest(-country, -topic) %>%
mutate(model=map(data, ~lm(percent_yes ~ year, data= .)),
tidied=map(model, tidy)) %>%
unnest(tidied)


# Print country_topic_coefficients
country_topic_coefficients

# Create country_topic_filtered
country_topic_filtered <- country_topic_coefficients %>%
filter(term=="year")

country_topic_filtered <- country_topic_filtered %>% mutate(p.adjusted=p.adjust(p.value)) %>%
filter(p.adjusted <0.05)

# Create vanuatu_by_country_year_topic
vanuatu_by_country_year_topic <- by_country_year_topic %>% 
filter (country=="Vanuatu")

# Plot of percentage "yes" over time, faceted by topic
ggplot(vanuatu_by_country_year_topic, aes(x=year, y=percent_yes)) + 
  geom_line() +
  facet_wrap (~topic)
```

